ASTR480 Final Report:\\ \textit{Asteroids in TESS}

### 2024-10-16 FinalReport/FinalReportDraft.tex
some homework

With TESS data coming in $\frac12\unit{\hour}$ chunks, 24 interpolated points are needed between each API call.
These interpolated positions can be seen in \autoref{Fig:interpPos} for a subsection of sector from \texttt{TESSELLATE}.
There are a few interesting features, such as the asteroids are moving in the same direction, indicated by the colouring, they come in from low RA and Dec and tend to increase both coordinates as the month of the sector progresses.
There is also a large size range in this slice of sky, ranging over \qty{5}{\mag} in absolute magnitude, which can be seen in the alpha (or transparency) of the tracks in \autoref{Fig:interpPos}.
Because of a small change in position between each frame, \citep[$\sim \qty{1}{\px}$ per frame][]{Pal2018}, this interpolation should be accurate.
The
Checking against a higher frequency query to the JPL Horizons ephemeris  confirmed this accuracy on a few asteroids, as shown in \autoref{Fig:errPos}.
This is not a feasible way of getting all the positions, as the queries are per name and time, not per area and time like the cone search in Skybot.
They are also rate limited, so while a few asteroids can be checked, filling the gaps between the Skybot half day queries for all the asteroids would take far too long.
For the faster TESS FFIs, more interpolated points are needed, but a smaller the change in position between each point will be.
Forcing the interpolations to be at the same time as the TESS frames, which are known thanks to \texttt{TESSELLATE} saving them out, simplifies the matching to detections and forced photometry of these points.

@@

The integer coordinates of the asteroids seemed to be leading the object in any given frame of TESS. %?FIG?
This was decreasing the total flux of the asteroid, as the box used for aperture  was hitting mostly background.
This also caused a ``sawtoothing'' effect, where the lightcurve would jump from a minimum to a maximum when the integer coordinate snapped from being a pixel out to back on the asteroid.
Both of these issues can be fixed by calculation of the centre of mass (COM), the average brightness in $xy$ space on a $5\times5$ \unit{\px} box of the image centred on the interpolated position.
I didn't take the highest pixel value in this box because there could be a rouge star in the field that it snaps to, instead of the asteroid.
A \texttt{Photutils} aperture photometry centred on the COM with the same radius of \qty{1.5}{\px} sorted some of the discrepancies in the lightcurve, with a far smoother transition from maximum to minimum, as expected from rotating asteroids. Hereafter, unless otherwise specified, ``flux'' refers to this COM flux measurement of the asteroid.

@@

There are a few different implementations of a Lomb-Scargle periodogram available in world of astronomy python packages.
I first tried using the \texttt{Lightkurve} \citep{Lightkurve2018} package built for period analysis of TESS (and Kepler) data of variable stars.
\texttt{Lightkurve} was easy to use but didn't allow for fine-tuning of the periodogram, or give any model of the curve it produced.
The next option was peeling back a layer of abstraction and using the Lomb-Scargle periodogram as implemented by \texttt{Astropy}\citep[\citet{Astropy2022} but see][for the implementation]{Vanderplas2012,Vanderplas2015}.
There were interesting similarities and differences between the packages.
\texttt{Lightkurve} is based on the \texttt{Astropy} methods, but does not give all the functionality, instead opting to simplify the process.
\texttt{Astropy} gives a statistic on how confident it is in the detected frequency is, the false alarm probability.
It also returns a model, in the form of

@@

\begin{figure}
  \centering
  \includegraphics[width=\textwidth]{../OzData/recoverdHistBkgLimof10AtanBothLegendFixed.pdf}
  \caption[Recovery plot]{A histogram of known average visual magnitude of each object against the whether the object has a lightcurve with an average flux value above 10 %!CHECK
  counts for its COM flux (blue triangles) or passes all of the checks on the lightcurve and periodogram (green dots).
  A fit of an $\arctan$ function is provided in dotted lines of the same colour. 
  }
  \label{Fig:RecovPassorFail}
\end{figure}

% \begin{figure}
%   \centering
%   \includegraphics[width = \textwidth]{../OzData/matchesHist29.pdf}
%   \caption[Magnitude of matches]{The visual magnitude of the asteroid against the fraction of asteroids in this bin with more than 3 %!CHECK
%     matches.
%   }
%   \label{Fig:MagofMatches}
% \end{figure}

\begin{figure} %TODO fix.
  \centering
  \includegraphics[width=\textwidth]{../OzData/pointsMatchesNumberHistdouble1.pdf}
  \caption[Number of observations or matches]{\textit{Left}: A histogram of the number of observations in the lightcurve of each asteroid (orange solid line) in 100 observation bins. \textit{Right}: The number of matches each asteroid gets in the \texttt{TESSELLATE} detections (black dashed)in 100 match bins.
  Both are limited to 600 asteroids on the y-axis, which pales in comparison to the number of asteroids with $\leq$ 100 matches and yet is too large for detail to be seen in any higher bins. }
  \label{Fig:MatchInterpHists}
\end{figure}

@@

@@

quality checks aren't great

run on rest of TESSELLATE data

>>>>

With TESS data coming in $\frac12\unit{\hour}$ chunks, 24 interpolated points are needed between each API call.
These interpolated positions can be seen in \autoref{Fig:interpPos} for a subsection of sector from \texttt{TESSELLATE}.
There are a few interesting features, such as the asteroids are moving in the same direction, indicated by the colouring, they come in from low RA and Dec and tend to increase both coordinates as the month of the sector progresses.
There is also a large size range in this slice of sky, ranging over \qty{5}{\mag} in absolute magnitude, which can be seen in the alpha (or transparency) of the tracks in \autoref{Fig:interpPos}.
Because of a small change in position between each frame, \citep[$\sim \qty{1}{\px}$ per frame][]{Pal2018}, this interpolation should be accurate.

Checking against a higher frequency query to the JPL Horizons ephemeris  confirmed this accuracy on a few asteroids, as shown in \autoref{Fig:errPos}.
This is not a feasible way of getting all the positions, as the queries are per name and time, not per area and time like the cone search in Skybot.
They are also rate limited, so while a few asteroids can be checked, filling the gaps between the Skybot half day queries for all the asteroids would take far too long.
For the faster TESS FFIs, more interpolated points are needed, but a smaller the change in position between each point will be.
Forcing the interpolations to be at the same time as the TESS frames, which are known thanks to \texttt{TESSELLATE} saving them out, simplifies the matching to detections and forced photometry of these points.

@@

The integer coordinates of the asteroids seemed to be leading the object in any given frame of TESS. %?FIG?
This was decreasing the total flux of the asteroid, as the box used for aperture  was hitting mostly background.
This also caused a ``sawtoothing'' effect, where the lightcurve would jump from a minimum to a maximum when the integer coordinate snapped from being a pixel out to back on the asteroid.
Both of these issues can be fixed by calculation of the centre of mass (COM), the average brightness in $xy$ space on a $5\times5$ \unit{\px} box of the image centred on the interpolated position.
I did not take the highest pixel value in this box because there could be a rouge star in the field that it snaps to, instead of the asteroid.
A \texttt{Photutils} aperture photometry centred on the COM with the same radius of \qty{1.5}{\px} sorted some of the discrepancies in the lightcurve, with a far smoother transition from maximum to minimum, as expected from rotating asteroids. Hereafter, unless otherwise specified, ``flux'' refers to this COM flux measurement of the asteroid.

@@

There are a few different implementations of a Lomb-Scargle periodogram available in world of astronomy python packages.
I first tried using the \texttt{Lightkurve} \citep{Lightkurve2018} package built for period analysis of TESS (and Kepler) data of variable stars.
\texttt{Lightkurve} was easy to use but did not allow for fine-tuning of the periodogram, or give any model of the curve it produced.
The next option was peeling back a layer of abstraction and using the Lomb-Scargle periodogram as implemented by \texttt{Astropy}\citep[\citet{Astropy2022} but see][for the implementation]{Vanderplas2012,Vanderplas2015}.
There were interesting similarities and differences between the packages.
\texttt{Lightkurve} is based on the \texttt{Astropy} methods, but does not give all the functionality, instead opting to simplify the process.
\texttt{Astropy} gives a statistic on how confident it is in the detected frequency is, the false alarm probability.
It also returns a model, in the form of

@@

\begin{figure} %? // TODO fix.
  \centering
  \includegraphics[width=\textwidth]{../OzData/pointsMatchesNumberHistdouble1.pdf}
  \caption[Number of observations or matches]{\textit{Left}: A histogram of the number of observations in the lightcurve of each asteroid (orange solid line) in 100 observation bins. \textit{Right}: The number of matches each asteroid gets in the \texttt{TESSELLATE} detections (black dashed).
  }
  \label{Fig:MatchInterpHists}
\end{figure}

The number of forced photometry observations in a lightcurve is more evenly distributed than the number of matches, as shown in \autoref{Fig:MatchInterpHists}.
The quality cut of 200 observations in a lightcurve only cuts out $\sim 200$ asteroids.
Most asteroids are well sampled, with the mode of the distribution in the 2900 to 3000 observation bin, which means the asteroids are in a large fraction of the about 3500 frames in a sector.

The matches have a very different distribution to the observations.
While both plots are limited to 600 asteroids on the y-axis, the matches histogram far exceeds this in the first bin, $\leq$ 100 matches, as indicated by the text and arrow.
Yet the y-limit is too large for detail to be seen in any higher bin.
There are only 102 asteroids matched more than 100 times in the \texttt{TESSELLATE} detections.

\begin{figure}
  \centering
  \includegraphics[width=\textwidth]{../OzData/recoverdHistBkgLimof10AtanBothLegendFixed.pdf}
  \caption[Recovery plot]{The recovery fraction against the known visual magnitude of each object against the whether the
    object exceeds only the averge flux requirement  of the quality checks (blue triangles) or passes all of the quality checks on the lightcurve and periodogram (green dots).
    A fit of an $\arctan$ function is provided in dotted lines of the same colour.
  }
  \label{Fig:RecovPassorFail}
\end{figure}

The fits of to the fraction recoverd in \autoref{Fig:RecovPassorFail} take the form
\begin{equation} \label{Eq:Atan}
  y= 0.5 -\frac{1}{\pi} \arctan(\alpha \cdot (m-m_{1/2}))
\end{equation}
where $y$ is the recoved fraction, $\alpha$ is how the steep the transition from 1 down to 0 is,  $m$ the magnitude and $m_{1/2}$ is the shift along the magnitude axis.
The constant of 0.5 in \autoref{Eq:Atan} raises the point of inflection to be at $\frac 1/2$, and the $\frac{-1}{\pi}$ makes the function take the expected maximum at bright magnitudes at \qty{100}{\percent} recoved, down to no dim asteroids being detected.
The $m_{1/2}$ parameter dictates the limiting magnitude, as dimmer than this value less than half of the asteroids are recoverd.
The fit is preformed on all bins (of \qty{0.5}{\mag}) dimmer than the first bin with more than one asteroid in it, to stop extremely bright asteroids from having too much influence.

Only taking the minimum average flux of 10 counts requirement, blue in \autoref{Fig:RecovPassorFail}, the limiting magnitude is \qty{17.78(0.01)}{\mag}.
When all the quality checks are imposed, this gets over a \unit{\mag} brighter, to \qty{16.46(0.37)}{\mag}.
It is expected that a higher threshold for quality will make the limiting magnitude brighter, but is value is so uncertain because of the scatter in the recovery fraction at brighter magnitudes, and the shallower decrease in efficiency.

@@

The limiting magnitude of the recovery fraction demonstrated in \autoref{Fig:RecovPassorFail} agrees well with other tests of \texttt{TESSELLATE}.
Concurrent work by another Honours student \citep{MontillaHons} achieves a similar recovery fraction detection efficiency of $\sim$\nth{17} \unit{\mag}.
She has been preforming injection/recovery tests across many sectors, and finding the limiting magnitude is consistent across sectors and cadence of the FFIs.
This is promising, as it means that asteroids will be detected to a similar depth as they are here for any further sectors processed, something I could not check myself.

@@

%quality checks aren't great
It is clear from the outliers in \autoref{Fig:perVSamp} and the clearly incorrect periods of know objects in \autoref{Fig:LCDBcomp} that the quality checks have room for improvement.
Some of these checks were  taken to agree with the how others analyse similar data, like the maximum period from \citet{McNeill2023}.
Others were more empherically derived, such as the minimum average flux value.
There is a delecate balance between detecting as many known periods as possible and maintating strict enough cuts to keep out the junk.

The periodogram noise at the low frequency end is quite strong throughout the range of rotation periods. 
Currently it is responsible for a lot of undetected rotaion rates, due to the \qty{90}{\percent} of the \qty{17}{\day} window check, and this noise region having a larger power than any signal.
Modelling this excess to try and remove the noise and boost the signal from the rotaion could be done to improve the maximum power of the periodogram. 
There are a few asteroids with known longer rotaion periods than this check allows for, but one TESS sector will not be enough to determine those accurately.

Linking lightcurves of asteroids between sectors was not considered here, as only one sector has been analysed.
Any one asteroid will be in multiple sectors, as TESS has surveyed that whole sky more than once. 
The different FFI cadence, as well as the large temporal gap between these observations makes it non-trivial to recover the rotation periods.
The different depths that the shortest cadence FFIs have \citep{MontillaHons}

%run on rest of TESSELLATE data
